Imperative Programming with Variational Effects

### 2018-07-01 paper/paper.tex
finish background section; work some on motivation

@@

If we do not have to worry about side effects in our language that we wish to introduce variation to
our task is comparatively easy. We begin by considering a simple arithmetic expression language:

$a ::= n\ |\ a + a\ |\ a * a$

TODO: demonstrate how this works for pure expressions

@@

In a non-variational context, the behavior of this program should be clear.
If the variable $y$ evaluates to \textbf{true}, then an error is thrown with value
$e$. At this point evaluation should stop, meaning that the variable $x$ is never
assigned in the final statement, avoiding the costly computation of \texttt{expensiveFn}. This effect of halting execution and
returning an error value is the essence of exceptions in non-variational settings.

Now we consider the behavior of the same program in a variational context.
Suppose that the variable $y$ evaluates to the variational value
$A \langle \textbf{true}, \textbf{false} \rangle$. This means that in the left alternative of $A$ we
would evaluate the body of the if statement, while ignoring it otherwise. Therefore,
in the left alternative of $A$ we throw an exception, but otherwise we continue on to the
final statement.

@@

We also can't simply continue evaluation
in every variant regardless of whether or not we have encountered an error.
Suppose that calling \texttt{expensiveFn} normally would evaluate to the variational value $A\langle 1, 2 \rangle$, but at considerable computational cost in both alternatives.
Because we know that the left alternative of $A$ will ultimately evaluate to the thrown exception $e$, we would
like to avoid the cost involved in computing the value 1 that we will just throw away later, mirroring how throwing an exception short-circuits evaluation in the
non-variational setting.

>>>>

\subsection{Implementation of Formula Trees}

When integrating formula trees into an existing non-variational language we can choose to
either embed choices directly into the abstract syntax of the host language or utilize a type-generic
implementation. In this work we employ both strategies, opting to embed choices directly when dealing
with programs, while using the type-generic implementation for the values produced by those programs.
Here we present the type-generic implementation, although the basic principles explored here apply
equally to embedded representations.

Generic formula trees can be represented by the following Haskell datatypes:

\begin{program}
type Var = String
data Dim =
    Lit Bool
  | Ref Var
  | Not Dim
  | And Dim Dim
  | Or Dim Dim

data FTree a = One a | Chc Dim (FTree a) (FTree a) 
\end{program}

The datatype \prog{Dim} represents expressions in our boolean algebra, with variable names as strings.
The constructor \prog{One} creates the leaves of a formula tree, while \prog{Chc} creates the nodes.

This type-generic representation allows us to easily define some useful algebraic properties of
formula trees via typeclasses. Specifically, formula trees readily support implementation of the
Functor, Applicative, and Monad typeclasses:

\begin{program}
instance Functor FTree where
  fmap f (One a) = One (f a)
  fmap f (Chc d l r) = Chc d (fmap f l) (fmap f r)

instance Applicative FTree where
  pure = One
  (One f) <*> v = fmap f v
  (Chc d l r) <*> v = Chc d (l <*> v) (r <*> v)

instance Monad FTree where
  return = One
  (One v) >>= f = f v
  (Chc d l r) >>= f = Chc d (l >>= f) (r >>= f)
\end{program}

Astute readers will note that because formula trees are merely a special class of binary tree with
labeled nodes they share identical instances. The usefulness of defining these instances will be made
apparent further on in this work during the discussion of pure, side-effect free variational languages.

@@

If we do not have to worry about side effects in our language that we wish to introduce variation to
our task is comparatively easy. We begin by considering a simple arithmetic expression language
defined by the following Haskell datatype:

\begin{program}
data Arith = N Int | Add Arith Arith | Mul Arith Arith
\end{program}

In the non-variational setting we can easily define an interpreter for this language which produces
values of type \prog{Int}:

\begin{program}
eval :: Arith -> Int
eval (N i) = i
eval (Add l r) = eval l + eval r
eval (Mul l r) = eval l * eval r
\end{program}

The process of ``variationalizing" this small example is quite straightforward. First, we add a
constructor to support choices in our language:

\begin{program}
data Arith = N Int | Add Arith Arith | Mul Arith Arith | AChc Dim Arith Arith
\end{program}

Next we modify our interpreter. Instead of returning plain values of type \prog{Int},
we now return \emph{variational} values of type \prog{FTree Int}. We then take the cases
from the non-variational interpreter and make them variational by situating them within the
applicative functor instance for \prog{FTree} we defined in Section 2. Finally, we add a case
that handles our new \prog{AChc} constructor and we have completed the conversion of our
interpreter:

\begin{program}
eval :: Arith -> FTree Int
eval (N i) = pure i
eval (Add l r) = (+) <$> eval l <*> eval r
eval (Mul l r) = (*) <$> eval l <*> eval r
eval (AChc d l r) = Chc d (eval l) (eval r)
\end{program}

This basic pattern of ``variationalizing" a language by re-situating the
plain interpreter within the applicative functor for choices works well for any language that is pure
and side-effect free. The only area of concern is associated with the use of data structures in the
interpretation of the language, which often benefit from being replaced with custom variational data structures that
provide greater sharing and efficiency. Recent work has explored the benefits of custom variational
data structures for maps \cite{erictodo}, linked lists \cite{karltodo}, and stacks \cite{mengtodo}.
As such, exercising proper care in the selection and integration of variational data structures
into a variational interpreter resolves this issue.

Nevertheless, even with the support of more efficient
variational data structures, our basic pattern of converting a plain interpreter to a variational one
encounters significant difficulties in the presence of a language with side effects.

@@

In a non-variational context, the behavior of this program should be clear.
If the variable $y$ evaluates to \textbf{true}, then an error is thrown with value
$e$. At this point evaluation should stop, meaning that the variable $x$ is never
assigned in the final statement, avoiding the costly computation of \prog{expensiveFn}. This effect of halting execution and
returning an error value is the essence of exceptions in non-variational settings.

Now we consider the behavior of the same program in a variational context.
Suppose that the variable $y$ evaluates to the variational value
$\tagtree{A}{true}{false}$. This means that in the left alternative of $A$ we
would evaluate the body of the if statement, while ignoring it otherwise. Therefore,
in the left alternative of $A$ we throw an exception, but otherwise we continue on to the
final statement.

@@

We also can't simply continue evaluation
in every variant regardless of whether or not we have encountered an error.
Suppose that calling \texttt{expensiveFn} normally would evaluate to the variational value $\tagtree{A}{1}{2}$, but at considerable computational cost in both alternatives.
Because we know that the left alternative of $A$ will ultimately evaluate to the thrown exception $e$, we would
like to avoid the cost involved in computing the value $1$ that we will just throw away later, mirroring how throwing an exception short-circuits evaluation in the
non-variational setting.
